Ensemble Stability of Biological Networks

### 2014-10-15 tex/figures.tex
recombine figures

\begin{figure}[!ht]
\centering
\noindent\includegraphics[width=1.0\columnwidth]{fig/combinedfigs.pdf}
\caption{{\bf Characterization of stability and robustness according to properties of system structure} (A) Stability versus connectivity, (B) robustness versus connectivity, (C) robustness versus number of cycles, (D) robustness versus hierarchy for three component systems.
}1,500 words long
\label{fig:combined}
\end{figure}

\begin{figure}[!ht]
\centering
\noindent\includegraphics[width=0.8\columnwidth]{fig/connectcycle3D3x3.pdf}
\caption{{\bf Robustness versus number of cycles and connectivity for three component systems.} Each point represents the average robustness of all graphs having a given number of cycles and connectivity.}
\label{fig:connectcycle3D3x3}
\end{figure}

\begin{figure}[!ht]
\centering
\noindent\includegraphics[width=0.8\columnwidth]{fig/connectdist3D3x3.pdf}
\caption{{\bf Robustness versus hierarchy and connectivity for three component systems.} Each point represents the average robustness of all graphs having a given hierarchy and connectivity.}
\label{fig:connectdist3D3x3}
\end{figure}

>>>>

\begin{figure}[!ht]
\centering
\noindent\includegraphics[width=1.0\columnwidth]{fig/combinedfigs.pdf}
\caption{{\bf Characterization of stability and robustness according to properties of system structure} (A) Robustness versus connectivity, (B) robustness versus hierarchy, (C) Robustness versus number of cycles and connectivity for three component systems. Each point represents the average robustness of all graphs having a given number of cycles and connectivity. (D) Robustness versus hierarchy and connectivity for three component systems. Each point represents the average robustness of all graphs having a given hierarchy and connectivity.
}
\label{fig:combined}
\end{figure}

% \begin{figure}[!ht]
% \centering
% \noindent\includegraphics[width=0.8\columnwidth]{fig/connectcycle3D3x3.pdf}
% \caption{{\bf Robustness versus number of cycles and connectivity for three component systems.} Each point represents the average robustness of all graphs having a given number of cycles and connectivity.}
% \label{fig:connectcycle3D3x3}
% \end{figure}

% \begin{figure}[!ht]
% \centering
% \noindent\includegraphics[width=0.8\columnwidth]{fig/connectdist3D3x3.pdf}
% \caption{{\bf Robustness versus hierarchy and connectivity for three component systems.} Each point represents the average robustness of all graphs having a given hierarchy and connectivity.}
% \label{fig:connectdist3D3x3}
% \end{figure}

### 2014-10-15 tex/supplement.tex
recombine figures

The analogous results for all three component systems are computed using Monte Carlo integration and shown in \ref{tab:structstabmat3} and \ref{fig:combined}B. This process is associated with some error relative to the exact integration described above. If we sample $N = N_{\mathrm{stab}} + N_{\mathrm{unstab}}$ matrices where each has some probability $\theta$ of being stable then $N_{\mathrm{stab}}$ has a binomial distribution. We can compute a sample estimate for $\theta$, $\hat{\theta} = \frac{N_{\mathrm{stab}}}{N}$ \cite{Murphy2012}. The posterior distribution in this case is known to be a Beta distribution as a result of Beta-Binomial conjugacy
$$
\mathrm{Beta}(\theta | \mathcal{D}) = \mathrm{Beta}(\theta | N_{\mathrm{stab}} + a, N_{\mathrm{unstab}} + b)
$$
where $a$ and $b$ are the hyperparameters of the Beta prior and we consider the uninformative uniform prior corresponding to $a=b=1$. We consider the maximum a posteriori estimate
$$\hat{\theta}_{MAP} = \frac{a + N_{\mathrm{stab}} - 1}{a + b + N - 2}$$
which corresponds in this case to the maximum likelihood estimate
$$
\hat{\theta}_{MLE} = \frac{N_{\mathrm{stab}}}{N}.
$$
This estimate is characterized by the variance of the posterior Beta distribution
$$
\mathrm{var}(\theta | \mathcal{D}) = \frac{(a+N_{\mathrm{stab}})(b+N_{\mathrm{unstab}})}{(a + N_{\mathrm{stab}} + b + N_{\mathrm{unstab}})^2(a + N_{\mathrm{stab}} + b + N_{\mathrm{unstab}}+1)}
$$
Since for the chosen prior $a=b=1 \ll N$ this simplifies to
$$
\mathrm{var}(\theta | \mathcal{D}) = \frac{\hat{\theta}(1-\hat{\theta})}{N}
$$
yielding the error estimate given by the associated standard deviation. In all simulations we use $N~=~10000$ so that the maximum error for $\hat{\theta}~=~0.5$ is $\sigma~=~\sqrt{\mathrm{var}(\theta | \mathcal{D})} \approx 0.005$.

It has been stated previously on the basis of simulation that system stability decreases with connectivity as the system size goes to infinity \cite{May1972}. For small system sizes such as the two and three component systems, studied here the situation is not so clear cut. For two component systems, system stability is constant across the entire range of connectivities. For three component systems, the trend shows a minor decrease from connectivity $4$ to $5$ followed by small fluctuations as shown in \ref{fig:combined}A.

The relationship between connectivity and robustness for two component systems is shown in \ref{tab:structstabmat} and likewise for three component systems in \ref{tab:structstabmat3} and \ref{fig:combined}B. If we average over the different classes of matrices for a given connectivity we see there is a correlation between connectivity and robustness demonstrated by the red dots in \ref{fig:combined}B.
% \subsection{Cycle number is inversely correlated with robustness}
\ref{fig:combined}C shows the robustness for all three component systems as a function of the number of simple cycles (elementary circuits) of length greater than one in the corresponding directed graph \cite{Johnson1975}. There appears to be a weak negative correlation between robustness and the number of simple cycles.

% \subsection{Cycle number and connectivity classify robust systems}
The combination of connectivity and cycle number as shown in \ref{fig:connectcycle3D3x3} provides a better classification of the dependence of robustness upon network topology. Here the robustness of three component systems with a given number of cycles, increases monotonically with connectivity. The network with the highest robustness for three component systems is that of \reffigscc (top panel). This network is the most hierarchical of all three component systems in the sense that it represents a total ordering of the components of the network and its adjacency matrix also shown in \reffigscc (top panel) has a block triangular structure.

This observation suggested that graph edit distance from \reffigscc (top panel), hierarchy, might provide a better characterization of system robustness. \ref{fig:combined}D shows system robustness as a function of hierarchy, where the most hierarchical system has a hierarchy of 0. There is a monotonic correlation between the upper bound of robustness and hierarchy. \ref{fig:connectdist3D3x3} shows system robustness as a function of both hierarchy and connectivity. The monotonic correlation between hierarchy and robustness is refined by an underlying correlation between robustness and connectivity analogous to that of \ref{fig:connectcycle3D3x3}.

>>>>

The analogous results for all three component systems are computed using Monte Carlo integration and shown in \ref{tab:structstabmat3} and \reffigrobustconnect. This process is associated with some error relative to the exact integration described above. If we sample $N = N_{\mathrm{stab}} + N_{\mathrm{unstab}}$ matrices where each has some probability $\theta$ of being stable then $N_{\mathrm{stab}}$ has a binomial distribution. We can compute a sample estimate for $\theta$, $\hat{\theta} = \frac{N_{\mathrm{stab}}}{N}$ \cite{Murphy2012}. The posterior distribution in this case is known to be a Beta distribution as a result of Beta-Binomial conjugacy
$$
\mathrm{Beta}(\theta | \mathcal{D}) = \mathrm{Beta}(\theta | N_{\mathrm{stab}} + a, N_{\mathrm{unstab}} + b)
$$
where $a$ and $b$ are the hyperparameters of the Beta prior and we consider the uninformative uniform prior corresponding to $a=b=1$. We consider the maximum a posteriori estimate
$$\hat{\theta}_{MAP} = \frac{a + N_{\mathrm{stab}} - 1}{a + b + N - 2}$$
which corresponds in this case to the maximum likelihood estimate
$$
\hat{\theta}_{MLE} = \frac{N_{\mathrm{stab}}}{N}.
$$
This estimate is characterized by the variance of the posterior Beta distribution
$$
\mathrm{var}(\theta | \mathcal{D}) = \frac{(a+N_{\mathrm{stab}})(b+N_{\mathrm{unstab}})}{(a + N_{\mathrm{stab}} + b + N_{\mathrm{unstab}})^2(a + N_{\mathrm{stab}} + b + N_{\mathrm{unstab}}+1)}
$$
Since for the chosen prior $a=b=1 \ll N$ this simplifies to
$$
\mathrm{var}(\theta | \mathcal{D}) = \frac{\hat{\theta}(1-\hat{\theta})}{N}
$$
yielding the error estimate given by the associated standard deviation. In all simulations we use $N~=~10000$ so that the maximum error for $\hat{\theta}~=~0.5$ is $\sigma~=~\sqrt{\mathrm{var}(\theta | \mathcal{D})} \approx 0.005$.

It has been stated previously on the basis of simulation that system stability decreases with connectivity as the system size goes to infinity \cite{May1972}. For small system sizes such as the two and three component systems, studied here the situation is not so clear cut. For two component systems, system stability is constant across the entire range of connectivities. For three component systems, the trend shows a minor decrease from connectivity $4$ to $5$ followed by small fluctuations as shown in \ref{fig:apstab3x3}.

The relationship between connectivity and robustness for two component systems is shown in \ref{tab:structstabmat} and likewise for three component systems in \ref{tab:structstabmat3} and \reffigrobustconnect. If we average over the different classes of matrices for a given connectivity we see there is a correlation between connectivity and robustness demonstrated by the red dots in \reffigrobustconnect.
% \subsection{Cycle number is inversely correlated with robustness}
\ref{fig:cycle3x3} shows the robustness for all three component systems as a function of the number of simple cycles (elementary circuits) of length greater than one in the corresponding directed graph \cite{Johnson1975}. There appears to be a weak negative correlation between robustness and the number of simple cycles.

% \subsection{Cycle number and connectivity classify robust systems}
The combination of connectivity and cycle number as shown in \reffigconnectcycle3D3x3 provides a better classification of the dependence of robustness upon network topology. Here the robustness of three component systems with a given number of cycles, increases monotonically with connectivity. The network with the highest robustness for three component systems is that of \reffigscc (top panel). This network is the most hierarchical of all three component systems in the sense that it represents a total ordering of the components of the network and its adjacency matrix also shown in \reffigscc (top panel) has a block triangular structure.

This observation suggested that graph edit distance from \reffigscc (top panel), hierarchy, might provide a better characterization of system robustness. \reffigrobusthierarchy shows system robustness as a function of hierarchy, where the most hierarchical system has a hierarchy of 0. There is a monotonic correlation between the upper bound of robustness and hierarchy. \reffigconnectdist3D3x3 shows system robustness as a function of both hierarchy and connectivity. The monotonic correlation between hierarchy and robustness is refined by an underlying correlation between robustness and connectivity analogous to that of \reffigconnectcycle3D3x3.
